Reproducible development environments with rix

### 2023-12-28 paper.qmd
typos and rewriting

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to list all the dependencies of the analysis
  clearly and ideally provide an easy way to install said dependencies.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will present
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that lowers
  Nix's learning curve for users of the [R]{.proglang} programming language.

@@

@peng2011 introduced the idea of reproducibility being on a continuum: on one of
the ends of this continuum, we only have access to the paper describing the
studies, which is not reproducible at all. Then, if in addition, to this paper
we make the original source code of the analysis that was written to compute the
results of the study available, reproducibility is improved, albeit only by a
little. Adding the original data improves reproducibility yet again. Finally, if
to all this we add what Roger Peng named the *linked and executable code and
data*, we reach the gold standard of full replication.

@@

Tools can be used at each of these steps to reach the gold standard of full
replication. Let's first consider the task of listing the software used.
[R]{.proglang} provides the `sessionInfo()` function whose output can be saved
into a file. Below is an example output of `sessionInfo()`:

@@

    ***and many more packages***
```

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific, library which doesn't interfere with the other,
main, library of the user. [renv]{.pkg} does not restore the [R]{.proglang}
itself though, so installing the right version of R needs to be handled
separately. Before continuing, it should be noted that other packages exist
which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
by @simonsohn2023 which makes it rather easy to install packages as they were
on CRAN at a given date. For example, the code snippet below install the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

@@

These packages also get installed in a project-specfic library so there is no
interferences between these packages and other versions of the same packages
that one might use for other projects. Because [groundhog]{.pkg} does not
install [R]{.proglang} itself, users should either install the required version
themselves, or they should use the `tolerate.R.version` argument as shown in the
example above. Otherwise, [groundhog]{.pkg} would not continue with the
installation of the packages. Another such package, developed by @chan2023 is
[rang]{.pkg}, which also installs packages as they were on a given date.
Yet another way to install packages as they were on a give date is to use
the Posit Package Manager, which provides snapshots of CRAN. For example,
to install the required packages for an analysis as they were on the 30th of June
2023, one could add the following line to the `.Rprofile` file:

@@

The `.Rprofile` file gets read by R when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror.

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process such
as [rig]{.proglang} by the R infrastructure team [-@rlib2023]. One could thus
install the right version of [R]{.proglang} using [rig]{.proglang} allows to
easily install different versions of [R]{.proglang}, one could use to install
the right version to reproduce an analysis, and then use one of the listed
packages above to install the right library of of [R]{.proglang} packages.
However, this involves many manual steps and is thus error prone. It is also
time-consuming to do so, especially if one wants to reproduce many studies.

The final step towards the gold standard of reproducibility is to package the
right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. From this image, containers can be run to reproduce
the data product with its exact dependencies. A statistical analysis, from the
simplest to the most complex one, can be seen as such a data product that comes
with many software dependencies. *Dockerizing* an analysis consists in first
building an image: in the build step, the dependencies of the analysis have to
be installed and this can be achieved using the tools mentioned above. The
scripts to run the analysis and the data are also added to the image at build
time. Instances of that image, called containers, can then be executed, which
give access to the exact development environment originally used for the study.
It is also possible to script the building of the analysis itself, such that in
one single call to the `docker run` command, a complete analysis is reproduced.

The Rocker project initiated by @boettiger2017 provides many pre-built
[Docker]{.proglang} to the [R]{.proglang} community of users. These images
can be used as bases to build other images containing statistical analyses
more easily than starting from a bare-bones image.

An optional step towards the gold standard is to use a build automation tool
such as [Make]{.proglang} to run the whole analysis when the container is
executed. Build automation tools make it easier to run arbitrary code in a
series of well-defined steps. [R]{.proglang} programmers can use the
[targets]{.pkg} by @landau2021 as a build automation tool.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.proglang} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses 
[Python]{.proglang} for statistical analysis instead.

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which available for all major operating system and focuses on
installing and building software in a reproducible manner. To make
[Nix]{.proglang} more accessible to [R]{.proglang} programmers, we developed the
[rix]{.pkg} which we will present in this article.

@@

[Nix]{.proglang} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, and the entirety of CRAN and Bioconductor is available
through [Nix]{.proglang}’s repositories. This means that using [Nix]{.proglang},
it is possible to install not only [R]{.proglang}, but also all the packages
required for a project. The reason why one should use [Nix]{.proglang} to
install [R]{.proglang} packages, and not use the usual, built-in,
`install.packages()` [R]{.proglang} function instead, is that when installing a
package with [Nix]{.proglang}, [Nix]{.proglang} makes sure to install every
dependency of every package, whether this dependency is another [R]{.proglang}
package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on the what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not easy, nor even possible. But with [Nix]{.proglang},
it suffices to declare that the [xlsx]{.pkg} package is needed for the project,
and [Nix]{.proglang} figures out automatically that [Java]{.proglang} is
required and installs and configures it. It all just happens without any
required intervention from the user.

But where do these packages come from? When installing a package using
[Nix]{.proglang}, an expression written in the [Nix]{.proglang} programming
language gets downloaded from the `nixpkgs` Github repository and evaluated.
This expression contains a so-called *derivation*. A derivation defines a build:
its dependencies, commands to build and install the package in question, and
then an output. Most of the time, a derivation downloads source code, builds the
software from the source and then outputs a compiled binary. Derivations are
extremely flexible, and by writing one's own, it is possible to define and build
environment for a project in a reproducible manner.

Because the whole set of [Nix]{.proglang} expression is hosted on Github, it is
possible to pin a specific revision of `nixpkgs` to ensure reproducibility of
our project. Pinning a revision ensures that every package that Nix installs
will always be at exactly the same versions, regardless of when in the future
the packages get installed.

With Nix, it is essentially possible to replace [renv]{.pkg} and Docker
combined, or in the case of [Python]{.proglang}, replace `requirements.txt`
files which list package dependencies for [Python]{.proglang} projects. It is
also possible to build multi-language environments, containing R and Python, a
LaTeX distribution, and even install one's favourite text editor to edit the
project's source code.

@@

Nix expressions are written in the [Nix]{.proglang} programming language,
which is a purely functional programming language. Here is a simple expression
that builds a shell environment containing version 4.3.1 of R:

@@

To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github respository, and in particular the commit starting
with `976fa336` is going to be used. It sould be noted that is thus possible to
use a personal fork of `nixpkgs` instead of the official repository.
`system_pkgs` lists the packages that we wish to install and use with this
environment, in this example only the [R]{.proglang} programming language will
be installed, alongside all the required dependecnies to actually install and
run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell
[R]{.proglang} will be started with the `--vanilla` flag (thus ignore any
startup options).

This expression can be saved in a file called `default.nix`, and the shell can
be built using the `nix-build` command. Once `nix-build` is done building the
development environment according to the specifications declared in the
`default.nix` file, the user can enter the development shell using `nix-shell`.
The user will thus be dropped into an interactive shell with all the packages
declared in the `default.nix` file and can now use this shell for work.

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite interesting. Because we believed that lowering the cost of
writing Nix expressions would lead to a wider adoption of the Nix package
manager for reproducibility, we wrote the [rix]{.pkg} which comes with the
`rix()` function that helps generate such expressions quite easily. For example,
here is how one could use [rix]{.pkg} to generate a `default.nix` file that
installs version 4.3.1 of [R]{.proglang} and the [dplyr]{.pkg} and
[chronicler]{.pkg} packages (we don't show the generated expression here because
of space):

@@

The above call to `rix()` generates an expression that installs several
[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
older version of the package. Had the user not specified the version, the
version available as of the `nixpkgs` commit would be used), but also install
several TeXLive packages for authoring documents with \LaTeX, the development
versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
the RStudio integrated development environment. It should be noted that as of
writing, RStudio is not available on macOS through `nixpkgs`, so the above
expression will only work on Windows and Linux distributions. To make the
expression work on macOS, the "rstudio" string could be replaced by "code",
which would install VS Code instead, or "other", to use any other editor
such as Emacs or Vi.

[rix]{.pkg} also provides installation intructions for Nix, but [rix]{.pkg} can
generate Nix expression even if Nix is not installed on one's system. This is
quite useful if one wants to generate an expression and use that expression in a
CI/CD environment such as Github Actions. For example, the repository that
contains the source code this article^[https://github.com/b-rodrigues/rix_paper]
uses Github Actions for compilation. Essentially, each time one of the authors
pushed to the master branch, a Github Actions runner would be started, Nix would
get installed on said runner, and an environment containing the right packages
was generated from a `default.nix` file that is hosted on the same repository.
The paper would then get compiled in this reproducible environment.

If one wishes to run pipelines written using the [targets]{.pkg} [@landau2021]
on Github Actions, [rix]{.pkg} featrues a function called `tar_nix_ga()`,
inspired by the `tar_github_actions()` function from the [targets]{.pkg}
package. This function writes a YAML file under `.github/workflows/` which then
runs the [targets]{.pkg} pipeline using Github Actions at each push, in a
completely reproducible environment. For such an example, consult the
code in this repository^[https://github.com/b-rodrigues/rix_cicd].

>>>>

--
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order to create an analysis that is easily reproducible,
  it is not enough to write clean code and document it well. One
  must also make sure to clear list all the dependencies of the analysis
  and ideally provide an easy way to install them.
  There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study,
  such as [Docker]{.proglang}, a containerization solution. This paper will
  present another tool that can be used to make projects reproducible,
  the Nix package manager, and an [R]{.proglang} package called [rix]{.pkg} that
  makes using Nix much simpler for users of [R]{.proglang}.

@@

@peng2011 introduced the idea of reproducibility being on a continuum: on one of
the ends of this continuum, we only have access to the paper describing the
studies, which is not reproducible at all. If in addition, to this paper authors
make the original source code of the analysis available, reproducibility is
improved, albeit only by a little. Adding the original data improves
reproducibility yet again. Finally, if to all this the authors add what Roger
Peng named the *linked and executable code and data*, we reach the gold standard
of reproducibility.

@@

There are tools available that authors can use to reach the gold stardard of a
fully reproducible study. Let us first consider the task of listing the software
used. [R]{.proglang} provides the `sessionInfo()` function whose output can be
saved into a file. Below is an example output of `sessionInfo()`:

@@

    ***lines below omitted***
```

This file lists every package alongside their versions and the repository from
which they were downloaded. Generating this file only requires one to run the
`renv::init()` function. Someone else can then restore the same package library
by running `renv::restore()`. The exact same packages get installed in an
isolated, project-specific, library which doesn't interfere with the main
library of the user. [renv]{.pkg} does not restore the [R]{.proglang} version
used though, so installing the right version of R needs to be handled
separately. Before continuing, it should be noted that other packages exist
which provide similar functionality to [renv]{.pkg}: there is [groundhog]{.pkg}
by @simonsohn2023 which makes it rather easy to install packages as they were on
CRAN at a given date. For example, the code snippet below installs the
[purrr]{.pkg} and [ggplot2]{.pkg} packages as they were on April 4th, 2017:

@@

These packages also get installed in a project-specfic library so there is no
interference between these packages and other versions of the same packages that
one might use for other projects. Because [groundhog]{.pkg} does not install
[R]{.proglang} itself, users should either install the required version
themselves, or they should use the `tolerate.R.version` argument as shown in the
example above. Otherwise, [groundhog]{.pkg} would not continue with the
installation of the packages. Another alternative to [renv]{.pkg}, developed by
@chan2023 is [rang]{.pkg}, which also installs packages as they were on a given
date. Yet another way to install packages as they were on a give date is to use
the Posit Package Manager, which provides snapshots of CRAN. For example, to
install the required packages for an analysis as they were on the 30th of June
2023, one could add the following line to the `.Rprofile` file:

@@

The `.Rprofile` file gets read by R when starting a new session, which means
that every call to the `install.packages()` function will now install the
packages from this snapshotted mirror. However, unless the user explicitely and
manually manages the different projects' libraries, using the Posit Package
Manager will install all the package for all the projects at the given date.

The next step in reaching the gold standard of reproducibility would be to not
only install the right packages used for the analysis, but also the right
version of [R]{.proglang}. Of course, it would be possible to install the right
version manually, but here too, there are tools that simplify the process. One
such tool is [rig]{.proglang} by the R infrastructure team [-@rlib2023].
[rig]{.proglang} makes it easy to install different versions of [R]{.proglang},
so one could use it to install the right version to reproduce an analysis, and
then use one of the listed packages above to install the right library of of
[R]{.proglang} packages. However, this involves many manual steps and is thus
error prone. It is also time-consuming to do so, especially if one wants to
reproduce many studies.

The final tool one could use to reach the gold standard of reproducibility is to
package the right version of [R]{.proglang} and [R]{.proglang} packages inside a
[Docker]{.proglang} image. [Docker]{.proglang} is a containerisation tool: using
[Docker]{.proglang} it is possible to package some *data product* with its
dependencies into an image. A statistical analysis, from the simplest to the
most complex one, can be seen as such a data product that comes with many
software dependencies. *Dockerizing* an analysis consists in first building an
image: in the build step, the dependencies of the analysis have to be installed
and this can be achieved using the tools mentioned above. The scripts to run the
analysis and the data are also added to the image at build time. Instances of
that image, called containers, can then be executed, which give access to the
exact development environment originally used for the study. This execution step
is called the "run step". During this run step, the analysis scripts can be
executed to product the output.

The advantage of using [Docker]{.proglang} is not only about having the right
version of [R]{.proglang} and [R]{.proglang} packages bundled alongside the
analysis scripts: [Docker]{.proglang} images are actually almost complete
operating systems, which means that system-level dependecies also get bundled
into the image. This is especially useful if packages that rely on external
libraries get used for the analysis, as these external libraries are thus also
made available to future reproducers of the study.

The Rocker project initiated by @boettiger2017 provides many pre-built
[Docker]{.proglang} to the [R]{.proglang} community of users, already containing
specific [R]{.proglang} versions and even packages. These images can be used as
bases to build other images containing statistical analyses more easily than
starting from a bare-bones image.

An optional step towards the gold standard is to use a build automation tool
such as [Make]{.proglang} to run the whole analysis when the container is
executed, instead of executing scripts. Build automation tools make it easier to
run arbitrary code in a series of well-defined steps. [R]{.proglang} programmers
can use the [targets]{.pkg} by @landau2021 as a build automation tool.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.proglang} for all the other
software, including the programming language that was used. Ideally, a build
automation tool such as [make]{.proglang} should be added to the list. It should
also be noted that these requirements are not limited to the [R]{.proglang}
programming language. A very similar approach should be taken if one uses
[Python]{.proglang} for statistical analysis instead. Furthermore, the previous
discussion also implies that for an analysis to actually be reproducible at all,
it must be made using free and open source software, as proprietary software
cannot be freely shared with others that wish to reproduce a study, and there's
no guarantee that others will have access to the proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.proglang}
package manager, which is available for all major operating systems and focuses
on installing and building software in a reproducible manner. This means that
this single tool can replace not only [Docker]{.proglang} but also [renv]{.pkg}
and a build automation tool such as [Make]{.proglang}. To make [Nix]{.proglang}
more accessible to [R]{.proglang} programmers, we developed the [rix]{.pkg}
package which we will present in this article.

@@

[Nix]{.proglang} is a package manager that can be used to install and build
software in a completely reproducible manner. As of writing, it contains more
than 80.000 packages, including the entirety of CRAN and Bioconductor. This
means that using [Nix]{.proglang}, it is possible to install not only
[R]{.proglang}, but also all the packages required for a project. The reason why
one should use [Nix]{.proglang} to install [R]{.proglang} packages, and not use
the usual, built-in, `install.packages()` [R]{.proglang} function instead, is
that when installing a package with [Nix]{.proglang}, [Nix]{.proglang} makes
sure to install every dependency of every package, whether this dependency is
another [R]{.proglang} package or a system-level dependency.

For example, the [xlsx]{.pkg} [R]{.proglang} package requires the
[Java]{.proglang} programming language to be installed to successfully install.
Depending on what system one tries to install [xlsx]{.pkg}, installing
[Java]{.proglang} might not be easy, nor even possible. But with
[Nix]{.proglang}, it suffices to declare that the [xlsx]{.pkg} package is needed
for the project, and [Nix]{.proglang} figures out automatically that
[Java]{.proglang} is required and installs and configures it. It all just
happens without any required intervention from the user.

But where do these packages come from? When installing a package using
[Nix]{.proglang}, an expression written in the [Nix]{.proglang} programming
language gets downloaded from the `nixpkgs` Github repository and evaluated.
This expression contains a so-called *derivation*. A derivation defines a build:
its dependencies, commands to build and install the package in question, and
then an output. Most of the time, a derivation downloads source code, builds the
software from the source and then outputs a compiled binary. Derivations are
extremely flexible, and by writing one's own, it is possible to define and build
an environment for a project in a reproducible manner.

But why is installing software with [Nix]{.proglang} reproducible? Because the
whole set of [Nix]{.proglang} expressions is hosted on Github, it is possible to
use a specific commit of `nixpkgs` to ensure reproducibility of our project.
Using a specific commit of `nixpkgs` is called *pinning a revision (or commit)*.
Pinning a revision ensures that every package that Nix installs will always be
at exactly the same versions, regardless of when in the future the packages get
built. This is because the [Nix]{.proglang} expressions that get downloaded when
pinning are the expressions as they were at that commit of `nixpkgs`.

With Nix, it is essentially possible to replace [renv]{.pkg} and
[Docker]{.proglang} combined, or in the case of [Python]{.proglang}, replace
`requirements.txt` files which list package dependencies for [Python]{.proglang}
projects. It is also possible to build multi-language environments, containing R
and Python, a LaTeX distribution, and even install one's favorite text editor to
edit the project's source code. Essentially, one can define a complete
reproducible environment and use it to work on a project interactively, or use
that environment non-interactively to run an analysis sometime in the future.

@@

[Nix]{.proglang} expressions are written in the [Nix]{.proglang} programming language,
which is a purely functional programming language. Here is a simple expression
that builds a shell environment containing version 4.3.1 of [R]{.proglang}:

@@

To define a variable in the [Nix]{.proglang}, the `let` keyword is used, so the
first lines of this expression define a variable called `pkgs` and
`system_pkgs`. `pkgs` is defined as the set of packages to be used in this
derivation. In this example, the package expressions are going to be pulled from
the official `nixpkgs` Github repository, and in particular the commit starting
with `976fa336` is going to be used. It should be noted that is also possible to
use a personal fork of `nixpkgs` instead of the official repository.
`system_pkgs` lists the packages that we wish to install and use with this
environment, in this example only the [R]{.proglang} programming language will
be installed, alongside all the required dependencies to actually install and
run [R]{.proglang}. Finally, the `mkShell` function is used to build a shell
containing the set of packages defined by the variable `system_packages` and its
`shellHook` is set to `R --vanilla`: this means that when entering the shell,
[R]{.proglang} will be started with the `--vanilla` flag (thus ignoring any
startup options).

This expression can be saved in a file called `default.nix`, and the shell can
be built using the `nix-build` command on a system where [Nix]{.proglang} is
available.^[To install [Nix]{.proglang} we recommend the installer from
Determinate systems which can be found on the following link:
\url{https://determinate.systems/posts/determinate-nix-installer}.] Once
`nix-build` is done building the development environment according to the
specifications declared in the `default.nix` file, the user can enter the
development shell using `nix-shell`. The user will thus be dropped into an
interactive shell with all the packages declared in the `default.nix` file and
can now use this shell for work. This is similar to activating environments in
the [Python]{.proglang} ecosystem.

Obviously, writing such an expression will be quite complicated for someone that
is unfamiliar with the [Nix]{.proglang} programming language, but the prospect
of being able to define a completely reproducible development environment in a
single text file and then use that file to build and use that environment
anywhere is quite appealing. Because we believe that lowering the cost of
writing [Nix]{.proglang} expressions would lead to a wider adoption of the
[Nix]{.proglang} package manager for reproducibility, we wrote [rix]{.pkg}
which comes with the `rix()` function that helps generate such expressions quite
easily. For example, here is how one could use [rix]{.pkg} to generate a
`default.nix` file that installs version 4.3.1 of [R]{.proglang} and the
[dplyr]{.pkg} and [chronicler]{.pkg} packages (we don't show the generated
expression here because of space):

@@

The above call to `rix()` generates an expression that installs several
[R]{.proglang} packages ([AER]{.pkg} version 1.2-8 gets installed, which is an
older version of the package. Had the user not specified the version, the
version available as of the `nixpkgs` commit would be used), but also install
several TeXLive packages for authoring documents with \LaTeX, the development
versions of [rix]{.pkg} and [fusen]{.pkg} (downloaded from Github), and finally
the RStudio integrated development environment. It should be noted that as of
writing, RStudio is not available on macOS through `nixpkgs`, so the above
expression will only work on Windows and Linux distributions. To make the
expression work on macOS, the "rstudio" string could be replaced by "code",
which would install VS Code instead, or "other", to use any other editor such as
Emacs or Vi. You can see the resulting `default.nix` file in the @sec-appendix.

[rix]{.pkg}'s documentation also provides installation instructions for Nix, but
[rix]{.pkg} can generate Nix expression even if Nix is not installed on one's
system. This is quite useful if one wants to generate an expression and use that
expression in a CI/CD environment such as Github Actions. For example, the
repository that contains the source code to this
article^[https://github.com/b-rodrigues/rix_paper] uses Github Actions for
compilation. Essentially, each time one of the authors pushed to the master
branch, a Github Actions runner would be started, [Nix]{.proglang} would get
installed on said runner, and an environment containing the right packages was
generated from a `default.nix` file that is hosted on the same repository. The
paper would then get compiled in this reproducible environment.

We also provide a way to start a temporary shell with [R]{.proglang} and
[rix]{.proglang} available, which makes it possible to start building
reproducible environments on a system that does provide [Nix]{.proglang} but not
[R]{.proglang}.

If one wishes to run pipelines written using the [targets]{.pkg} package
[@landau2021] on Github Actions, [rix]{.pkg} features a function called
`tar_nix_ga()`, inspired by the `tar_github_actions()` function from the
[targets]{.pkg} package. This function writes a YAML file under
`.github/workflows/` which then runs the [targets]{.pkg} pipeline using Github
Actions at each push, in a completely reproducible environment. For such an
example, consult the code in this
repository^[https://github.com/b-rodrigues/rix_cicd].
